A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-31 tidyindex.tex
knit

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

@@

Indexes are commonly used to combine multivariate information into a
single number for monitoring, communicating, and decision-making across
various domains in society. Their applications can be found in the
environment (e.g.~Air Quality Index, El Niño-Southern Oscillation
Index), in the economy (Consumer Price Index, stock market indexes), and
in the social sciences (e.g.~QS University Rankings, Human Development
Index).

To construct an index, experts typically first define an underlying
concept of interest to measure. This concept is latent and often without
direct measure but of social and public importance. Relevant and
available variables are then defined, collected, and combined as an
index using statistical methods to approximate the concept. This
construction process is often not straight forward and many decisions
need to be made, whether explicitly or not. For example, decision on
which variables to include may depend on the data availability and the
choice in other similar indexes. Indexes constructed from linear
combination of variables need to decide on the weight assigned to each
variable. Some indexes have a spatial and/or temporal component and
variables can be aggregated to different spatial resolutions and
temporal scales for different monitoring purpose. These decisions can
result in different index values and have different practical
implication.

To be able to test different decision choices systematically for an
index, the index needs to be broken down into its fundamental building
blocks to analyse the effect of each component. Such decomposition of
indexes standardise the index construction pipeline and offers benefit
for comparing among indexes and calculate index uncertainty. In social
indexes, the OECD handbook (OECD, European Union, and Joint Research
Centre - European Commission 2008) has provided a set of steps for
constructing composite socio-economic indexes, but there is a need to
extend these guidelines to accommodate more complex construction steps
required for indexes in general.

In this work, we develop a data pipeline framework where indexes from
different domains can be built from. Based on the pipeline, an R
package, tidyindex, is developed for practitioners to construct indexes
and understand how an index responds to changes in different components
in the pipeline. This work provides researchers actively developing new
indexes with tools to evaluate the proposed indexes on their robustness
for wider adoption. It also helps index analysts to diagnose an index to
identify its weakness for methodology improvement.

The rest of the paper is structured as follows:
Section~\ref{sec-pipeline} reviews the concept of data pipeline in R.
The pipeline framework for index construction is presented in
Section~\ref{sec-index-pipeline}. Section~\ref{sec-dev} explains how to
include a new building block in each pipeline module. Examples are given
in Section~\ref{sec-examples} to demonstrate the index construction with
the pipeline built.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, frame hidden, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Indexes are commonly used to combine and summarize different sources of
information into a single number for monitoring, communicating, and
decision-making. Examples of those include the Air Quality Index, El
Niño-Southern Oscillation Index, Consumer Price Index, QS University
Rankings or Human Development Index among many others.

To construct an index, experts typically start by defining a concept of
interest that requires measurement. This concept often lacks a direct
measurable attribute or can only be measure as a composite of various
processes, yet it holds social and public significance. To create an
index, once the underlying processes involved are identified, relevant
and available variables are then defined, collected, and combined using
statistical methods into an index that aims at measure the process of
interest. The construction process is often not straightforward, and
decisions need to be made, such as the selection of variables to be
included, which might depend on data availability and the statistical
definition of the index to be used, among others. For instance, the
indexes constructed from linear combination of variables need to decide
on the weight assigned to each variable. Some indexes have a spatial
and/or temporal components, and variables can be aggregated to different
spatial resolutions and temporal scales, leading to various indexes for
different monitoring purposes. Hence, all these decisions can result in
different index values and have different practical implications.

To be able to test different decision choices systematically for an
index, the index needs to be broken down into its fundamental building
blocks to analyse the contribution and effect of each component. We call
this process of breaking the index construction into different steps the
index pipeline. Such decomposition of index components provides the
means to standardise index construction via a pipeline and offers
benefits for comparing among indexes and calculating index uncertainty.
In social indexes for instance, the OECD handbook (OECD, European Union,
and Joint Research Centre - European Commission 2008) has provided a set
of steps and recommendations for constructing composite socio-economic
indexes. However, there is still a need to extend these guidelines and
methods to accommodate the inclusion of more methodological complex
steps required for indexes in general.

In this work, we aim at providing statistical and computational methods
to develop a data pipeline framework for constructing and customize
indexes using data. As a companion the R package, tidyindex, is
developed to construct and explore different. Furthermore, the tidyindex
package can be used to explore the effects of changes on index
definition, index construction steps, and construction methods. This
work provides researchers actively developing new indexes or aiming at
improving indexes with the tools to easily modify and evaluate indexes.
It also helps index analysts diagnose indexes, carry out sensitivity
analysis to assess small perturbations in the index, and identify
potential weaknesses for methodology improvement.

The rest of the paper is structured as follows:
Section~\ref{sec-pipeline} reviews the concept of data pipeline in R.
The pipeline framework for index construction is presented in
Section~\ref{sec-index-pipeline}. Section~\ref{sec-dev} explains how to
add new steps into the index methodology in the form of new building
blocks inside the index construction pipeline. Examples are given in
Section~\ref{sec-examples} to demonstrate the index construction with
the pipeline built.
